---
title: "Final Project Part 2"
author: "J Thomas"
date: "2024-12-10"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Step 1 - Identify and Import the data

I will start by reading in the Johns Hopkins data from the four .csv files.

```{r get_jhu_data}
# Get data from four files
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/csse_covid_19_time_series/"

file_names <- c("time_series_covid19_confirmed_US.csv", "time_series_covid19_confirmed_global.csv", 
"time_series_covid19_deaths_US.csv", 
"time_series_covid19_deaths_global.csv")

urls <- str_c(url_in, file_names)
```

Read in the datasets

```{r import_data}
us_cases <- read_csv(urls[1])
global_cases <- read_csv(urls[2])
us_deaths <- read_csv(urls[3])
global_deaths <- read_csv(urls[4])
```
Read in the data and transform (remove Lat and Long and rename Region and State.)

```{r tidy_global_data}
global_cases <- global_cases %>% 
  pivot_longer(cols=-c(`Province/State`, 
                        `Country/Region`, Lat, Long),
                names_to = "date",
                values_to = "cases") %>%
  select(-c(Lat, Long))

global_deaths <- global_deaths %>% 
  pivot_longer(cols=-c(`Province/State`, 
                        `Country/Region`, Lat, Long),
                names_to = "date",
                values_to = "deaths") %>%
  select(-c(Lat, Long))
```
Next, begin filtering and aggregating data.
```{r aggregate_global_data}
global <- global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))

global <- global %>% filter(cases > 0)

summary(global)
```
Clean US data
```{r tidy_us_data}
us_cases <- us_cases %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "cases") %>%
  select(Admin2:cases) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))

us_deaths <- us_deaths %>%
  pivot_longer(cols = -(UID:Combined_Key),
               names_to = "date",
               values_to = "deaths") %>%
  select(Admin2:deaths) %>%
  mutate(date = mdy(date)) %>%
  select(-c(Lat, Long_))
```
Join US datasets
```{r join_us_data}
us <- us_cases %>%
  full_join(us_deaths)
```
Combine US and Global Datasets
```{r combine_datasets}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)
```
Add population data
```{r add_population_data}
uid_lookup_url <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/refs/heads/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read_csv(uid_lookup_url) %>%
  select(-c(Lat, Long_, Combined_Key, code3, iso2, iso3, Admin2))
global <- global %>%
  left_join(uid, by = c("Province_State", "Country_Region")) %>%
  select(-c(UID, FIPS)) %>%
  select(Province_State, Country_Region, date, cases, deaths, Population, Combined_Key)
```
Visualize Data and Do Analysis
```{r us_by_state}
us_by_state <- us %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths)) %>%
  select(Province_State, Country_Region, date, cases, deaths) %>%
  ungroup()
us_by_state
```
```{r us_totals}
us_totals <- us %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths)) %>%
  select(Country_Region, date, cases, deaths) %>%
  ungroup()
us_totals
```
```{r visualize_us}
us_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID 19 in the US", y = NULL)
```
```{r visualize_co}
state = "Colorado"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) + 
  geom_point(aes(y = deaths, color = "deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID 19 in Colorado", y = NULL)
```
Transform State Data
```{r visualize_co}
us_by_state <- us_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
us_totals <- us_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```
```{r visualize_us_new_cases}
us_totals %>%
  filter(new_cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID 19 in the US", y = NULL)
```
```{r visualize_co_new_cases}
state = "Colorado"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(new_cases > 0) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) + 
  geom_point(aes(y = new_deaths, color = "new_deaths")) +
  scale_y_log10() + 
  theme(legend.position="bottom",
        axis.text.x = element_text(angle = 90)) + 
  labs(title = "COVID 19 in Colorado", y = NULL)
```
US State Totals
```{r us_state_death_toll}
us_state_max_deaths <- us_by_state %>%
  group_by(Province_State) %>%
  summarize(deaths = max(deaths))
us_state_max_deaths
```
Modeling
```{r create_model}
us_by_state <- na.omit(us_by_state)
mod <- lm(deaths ~ cases, data=us_by_state)
summary(mod)
us_by_state <- us_by_state %>% mutate(pred = predict(mod))
state = "Colorado"
us_by_state %>%
  filter(Province_State == state) %>%
  ggplot() +
  geom_point(aes(x=cases, y=deaths), color="blue") +
  geom_point(aes(x=cases, y=pred), color = "red")
```